---
title: "CES-D"
author: "Ana Luísa Marlière Casela; Leonardo Fernandes Martins; Pollyanna Santos da Silveira; Kimber Richter; Telmo Mota Ronzani"
date: "03 de fevereiro de 2016"
output:
  html_document:
    fig_caption: yes
    fig_height: 7
    fig_width: 9
    keep_md: yes
    toc: yes
  pdf_document:
    toc: yes
---

This is a public document with all scripts used to review the manuscript: 

"Psychometric properties of the Center for Epidemiologic Studies Depression Scale (CES-D) among people in treatment for substance use disorders"

All files used here are availible in a public repository accessible by the following url:

https://github.com/crepeia/CES-D_SubstanceAbuse


#Abstract
ABSTRACT
Objective: Depression is an important and prevalent comorbidity for substance addiction affecting the patient’s quality of life and decreasing adherence to substance abuse treatment. Measuring depressive symptoms is considered a crucial step for treatment. Among the instruments developed that measure the level of self-reported depression symptoms, the Center for Epidemiologic Studies Depression Scale (CES-D) has been widely used in clinical and populational studies. The present study aimed to evaluate psychometric properties of CES-D for a clinical sample of persons with drug addiction diagnostic. Method: 513 outpatients from two public health services for substance abuse treatment participated of this study. We evaluated construct validity in our sample using principal components analysis (PCA) and a confirmatory factorial analysis (CFA) to compare different models. Divergent and convergent validity was evaluated using Pearson correlation test. Reliability coefficients were calculated  by Cronbach’s alpha (α) and average split-half test. Results: PCA indicated a two dimensional structure explaining 47% of total variance. Based on these components we proposed a two factor structure that had a good fit for our sample  (χ2(116)=176.935, p<0.001; CFI=0.995; TLI=0.994; RMSEA=0.032,). The first factor (F1) composed by negative affects and depression symptoms items had high internal consistency (α=0.88) and a second one (F2) measuring absence of positive affect had low/moderated internal consistency (α=0.67). Convergent validity was supported by moderate positive correlation between CES-D factors and Self-Stigma (F1, r=0.45, p<0.001; F2, r=0.23, p<0,001). Divergent validity was supported by negative moderated correlation with Hope (F1, r = -0.33, p< 0.001; F2, r= -0.39; p<0.001) and Self-Esteem (F1, r = 0.45, p< 0.001; F2, r= 0.24, p<0.001). Our model showed a slightly better performance compared to other structures proposed by previous Brazilian studies. Conclusion: The psychometric properties of our proposed model, considering validity and reliability measures supports the use of CES-D to evaluate depression symptoms among people in treatment for substance use disorders. Keywords: Depression; Psychometrics; Validation Studies and Substance-Related Disorders.



#Preparing new analysis to CES-D manuscript review

Loading required packages
```{r}
require(foreign) # Read data stored SPSS
require(car) #Recode Variables
require(psych) #Psychometrics
require(lavaan) #Confirmatory and SEM
require(semPlot) # Plots for SEM
require(semTools) # Comparing SEM models
```

```{r}
#Setting Directory
setwd("~/CES-D_SubstanceAbuse")

#Importing SPSS file .sav
base.dat <- read.spss("PD10.sav", to.data.frame = T)

#Sum CESD itens in order to find NA
base.dat$scaleSum <- rowSums(base.dat[,267:286])

#Recoding reversed itens
base.dat$F4r<- recode(base.dat$F4r, "0=3; 1=2; 2=1; 3=0")
base.dat$F8r<- recode(base.dat$F8r, "0=3; 1=2; 2=1; 3=0")
base.dat$F12r<- recode(base.dat$F12r, "0=3; 1=2; 2=1; 3=0")
base.dat$F16r<- recode(base.dat$F16r, "0=3; 1=2; 2=1; 3=0")

#Creating a subset for analysis without NA
base.CESD <- subset(base.dat, subset=!is.na(base.dat$scaleSum))

#Creating a subset only with CESD
fullScale  <- base.CESD[ , 267:286]

#Creating a polychoric correlation
fullScaleT<-polychoric(fullScale)

#Cloning fullScale
fullScale2 <- fullScale

#Creating a categorical ordered subset for CFA analisys
orderedScale <-fullScale2[,c("F1r",
        "F2r",
        "F3r",
        "F4r",
        "F5r",
        "F6r",
        "F7r",
        "F8r",
        "F9r",
        "F10r",
        "F11r",
        "F12r",
        "F13r",
        "F14r",
        "F15r",
        "F16r",
        "F17r",
        "F18r",
        "F19r",
        "F20r")] <-
lapply(fullScale2[,c("F1r",
        "F2r",
        "F3r",
        "F4r",
        "F5r",
        "F6r",
        "F7r",
        "F8r",
        "F9r",
        "F10r",
        "F11r",
        "F12r",
        "F13r",
        "F14r",
        "F15r",
        "F16r",
        "F17r",
        "F18r",
        "F19r",
        "F20r")], ordered)
orderedScale<-as.data.frame(orderedScale)
```

```{r}
# Bartlett Test
cortest.bartlett(fullScaleT$rho, n=nrow(fullScale))
```

```{r}
# KMO
KMO(fullScaleT$rho)
```

```{r}
# Setting a randon number generator
set.seed(234) 
# Parallel Analysis
fa.parallel(fullScaleT$rho, fm="minres", fa="pc", n.obs=513, main = "Figure 1. Parallel Analysis Scree Plots")
```

#Principal Components Analysis
```{r}
#PCA - 2 components oblique rotated (assuming the components are correlated)
PCA2 <- principal(fullScaleT$rho, nfactors = 2, rotate="oblimin", method = "minres")
print.psych(PCA2, digits=2, cut= .4)
plot.psych(PCA2, title="Figure 2. Principal Components Analysis (CES-D)", title(sub = "correlation matrix=polychoric; method=minimal residual solution; rotation=oblimin", font.sub = 1, cex.sub = 0.5))

#Component 1
C1_PCA2 <- fullScale[, c("F1r","F2r","F3r","F5r","F6r","F7r","F9r","F10r","F11r","F13r","F14r","F15r","F17r","F18r","F19r","F20r")]
alpha(C1_PCA2, check.keys = TRUE)
splitHalf(C1_PCA2)

#Component 2
C2_PCA2 <- fullScale[, c("F4r","F8r","F12r","F16r")]
alpha(C2_PCA2, check.keys = TRUE)
splitHalf(C2_PCA2)
```

#Confirmatory Models

```{r}
#Batistoni CFA Model

#Model Identification
Batistoni <- '
              # latent variable definitions 
               f1 =~ F18r + F14r + F6r + F13r + F19r + F17r + F9r + F10r + F20r
               f2 =~ F3r + F1r + F7r + F5r + F2r
               f3 =~F12r + F8r + F16r + F4r

              # variances and covariances 
               f1 ~~ f2 
               f2 ~~ f3 
               f1 ~~ f3
                        '
#CFA Fit
fitBatistoni <- cfa(Batistoni, estimator="WLSMVS", mimic = "Mplus", data = orderedScale,
           ordered=c("F1r",
        "F2r",
        "F3r",
        "F4r",
        "F5r",
        "F6r",
        "F7r",
        "F8r",
        "F9r",
        "F10r",
        "F11r",
        "F12r",
        "F13r",
        "F14r",
        "F15r",
        "F16r",
        "F17r",
        "F18r",
        "F19r",
        "F20r"))

#Model Summary 
summary(fitBatistoni, standardized=T, fit.measures=T, rsquare=T)

#Model Fit Measures
fitMeasures(fitBatistoni, c("chisq","df","rmsea","rmsea.ci.lower", "rmsea.ci.upper", "srmr", "cfi", "tli", "nfi", "ecvi"))


#Modification Index
MIBatistoni<-modindices(fitBatistoni)
MIIBatistoni<- MIBatistoni[which(MIBatistoni$mi>30),]
print(MIIBatistoni)
```


```{r}
#Silveira CFA Model

#Model Identification
Silveira <- '
              # latent variable definitions 
               f1 =~ F18r + F14r + F6r + F3r + F13r
               f2 =~ F19r + F15r + F17r + F1r + F9r + F10r
               f3 =~ F20r + F7r + F5r + F11r
               f4 =~ F12r + F8r + F16r

             # variances and covariances 
               f1 ~~ f2 
               f2 ~~ f3 
               f3 ~~ f4
               f1 ~~ f4
               f1 ~~ f3 
                       '

#Model Fit
fitSilveira <- cfa(Silveira, estimator="WLSMVS", mimic = "Mplus", data = orderedScale,
        ordered=c("F1r",
        "F2r",
        "F3r",
        "F4r",
        "F5r",
        "F6r",
        "F7r",
        "F8r",
        "F9r",
        "F10r",
        "F11r",
        "F12r",
        "F13r",
        "F14r",
        "F15r",
        "F16r",
        "F17r",
        "F18r",
        "F19r",
        "F20r"))

#Model Summary 
summary(fitSilveira, standardized=T, fit.measures=T, rsquare=T)

#Model Fit Measures
fitMeasures(fitSilveira, c("chisq","df","rmsea","rmsea.ci.lower", "rmsea.ci.upper", "srmr", "cfi", "tli", "nfi", "ecvi"))

#Modification Index
MISilveira<-modindices(fitSilveira)
MIISilveira<- MISilveira[which(MISilveira$mi>30),]
print(MIISilveira)
```


```{r}
#Marliere - Principal Components Analysis - Two Components Solution - CFA Model 

PCA2_CFA <- '
              # latent variable definitions 
               f1 =~ F4r + F8r + F12r + F16r
               f2 =~ F1r + F2r + F3r + F5r + F6r + F7r + F9r + F10r + F11r + F13r + F14r + F15r + F17r + F18r + F19r + F20r

                #factor covariances 
                f1~~f2
                       '
fitPCA2 <- cfa(PCA2_CFA, estimator="WLSMVS", mimic = "Mplus", data = orderedScale,
        ordered=c("F1r",
        "F2r",
        "F3r",
        "F4r",
        "F5r",
        "F6r",
        "F7r",
        "F8r",
        "F9r",
        "F10r",
        "F11r",
        "F12r",
        "F13r",
        "F14r",
        "F15r",
        "F16r",
        "F17r",
        "F18r",
        "F19r",
        "F20r"))

#Model Summary 
summary(fitPCA2, standardized=T, fit.measures=T, rsquare=T)

#Model Fit Measures
fitMeasures(fitPCA2, c("chisq","df","rmsea","rmsea.ci.lower", "rmsea.ci.upper", "srmr", "cfi", "tli", "nfi", "ecvi"))

#Parameters Estimates
EstPCA2 <- parameterEstimates(fitPCA2, standardized=T, ci=F)
subset(EstPCA2, op == "=~")

#Modification Index
MIPCA2<-modindices(fitPCA2)
MIIPCA2<- MIPCA2[which(MIPCA2$mi>30),]
print(MIIPCA2)
```


```{r}
#Marliere - Final Solution - Principal Components Analysis - Two Components Solution Reviewed - CFA Model (removed itens - F4r, F8r, F13r) - and included error covariance

PCA2_CFArf <- '
              # latent variable definitions 
               f2 =~ F12r + F16r
               f1 =~ F1r + F2r + F3r + F5r + F6r + F7r + F9r + F10r + F11r + F14r + F15r + F17r + F18r + F19r + F20r
              #factor covariances 
                f1 ~~ f2    
              #error covariance
                F15r ~~ F19r
                F17r ~~ F18r
                       '
fitPCA2rf <- sem(PCA2_CFArf, estimator="WLSMVS", mimic = "Mplus", data = orderedScale,
        ordered=c("F1r",
        "F2r",
        "F3r",
        "F4r",
        "F5r",
        "F6r",
        "F7r",
        "F8r",
        "F9r",
        "F10r",
        "F11r",
        "F12r",
        "F13r",
        "F14r",
        "F15r",
        "F16r",
        "F17r",
        "F18r",
        "F19r",
        "F20r"))

#Model Summary 
summary(fitPCA2rf, standardized=T, fit.measures=T, rsquare=T)

#Model Fit Measures
fitMeasures(fitPCA2rf, c("chisq","df","rmsea","rmsea.ci.lower", "rmsea.ci.upper", "srmr", "cfi", "tli", "nfi", "ecvi"))


#Parameters Estimates
EstPCA2rf <- parameterEstimates(fitPCA2rf, standardized=T, ci=F)
subset(EstPCA2rf, op == "=~")

#Modification Index
MIPCA2rf<-modindices(fitPCA2rf)
MIIPCA2rf<- MIPCA2rf[which(MIPCA2rf$mi>30),]
print(MIIPCA2rf)

#Component 1
C1_PCA2final <- fullScale[, c("F1r","F2r","F3r","F5r","F6r","F7r","F9r","F10r","F11r","F14r","F15r","F17r","F18r","F19r","F20r")]
alpha(C1_PCA2final, check.keys = TRUE)
splitHalf(C1_PCA2final)

#Component 2
C2_PCA2final <- fullScale[, c("F12r","F16r")]
alpha(C2_PCA2final, check.keys = TRUE)
splitHalf(C2_PCA2final)

#Model Plot
semPaths(fitPCA2rf, what="path", whatLabels ="std", edge.label.cex = 0.7, exoVar = F, exoCov = T, layout = "tree2", optimizeLatRes=T, style = "lisrel", curve= 0.9, sizeLat = 5, sizeLat2 = 5, sizeMan = 3, sizeMan2 = 3, title = T, thresholds = F, curvePivot=T, intercepts = F, residuals = T)

#Define Title
title(main = "Figure 3. CES-D - Final Model for Confirmatory Factorial Analysis", line = 1)

#Define Subtitle
title(sub = expression("Fit measures:" ~ chi^2~(116)==176.935 ~", p<0.001, n=512; CFI=0.995; TLI=0.994; NFI=0.986; RMSEA=0.032, 90%CI(0.022-0.041); SRMR=0.045"), line = 3, font.sub = 1, cex.sub = 0.5)
```


#Comparing Fits
```{r}
fit<-compareFit(fitPCA2rf,fitPCA2,fitBatistoni,fitSilveira, nested = F)
print(fit)
summary(fit, fit.measures="all")
```

#More Fits
```{r}
moreFitIndices(fitPCA2rf, fit.measures = "all", nPrior = 513)
moreFitIndices(fitPCA2, fit.measures = "all", nPrior = 513)
moreFitIndices(fitBatistoni, fit.measures = "all", nPrior = 513)
moreFitIndices(fitSilveira, fit.measures = "all", nPrior = 513)
```

#Final Solution
```{r}
#CFA2
#Sum CESD itens PCA2

#Component 1
base.CESD$PCA2C1 <- base.CESD$F1r+ base.CESD$F2r+ base.CESD$F3r+ base.CESD$F5r+ base.CESD$F6r+ base.CESD$F7r+ base.CESD$F9r+ base.CESD$F10r+ base.CESD$F11r+ base.CESD$F14r+ base.CESD$F15r+ base.CESD$F17r+ base.CESD$F18r+ base.CESD$F19r+ base.CESD$F20r

#Component 2
base.CESD$PCA2C2 <- base.CESD$F12r+ base.CESD$F16r
```

```{r}
#Correlation - ISMI and CES-D

#Correlation ISMI Total Socre x CESD
PCA2xISMIgeral<- base.CESD[,c(230:231, 288:289)]
corr.test(PCA2xISMIgeral,  use="pairwise.complete.obs")

#Correlation ISMI Fatores x CESD
PCA2xISMIfatores<- base.CESD[,c(232:236, 288:289)]
corr.test(PCA2xISMIfatores,  use="pairwise.complete.obs")

#Correlation Hope x CESD
PCA2xEE<- base.CESD[,c(241,288:289)]
corr.test(PCA2xEE,  use="pairwise.complete.obs")

##Correlation Self-Esteem x CESD
PCA2xEAUE<-base.CESD[,c(253,288:289)]
corr.test(PCA2xEAUE,  use="pairwise.complete.obs")
```

```{r}
#Summaries for PCA2
summary(base.CESD$PCA2C1)
sd(base.CESD$PCA2C1)
summary(base.CESD$PCA2C2)
sd(base.CESD$PCA2C2)
#Plots for each scale
par(mfrow=c(2,2))
hist(base.CESD$PCA2C1, breaks=10, axes=F, main="", xlab="", ylab="")
axis(2)
hist(base.CESD$PCA2C2, breaks=5,  axes=F, main="", xlab="", ylab="")
axis(2)
boxplot(base.CESD$PCA2C1, horizontal=T, axes=F)
axis(1)
mtext("CESD - Subscale Factor 1 (M=25.0; SD=11.04)", side=1, line=2.5, font=2)
boxplot(base.CESD$PCA2C2, horizontal=T, axes=F)
axis(1)
mtext("CESD - Subscale Factor 2 (M=3.1; SD=2.17)", side=1, line=2.5, font=2)
title("Figure 4. Histogram and Boxplot for CES-D Final Solution (n=513) ", outer=T, line=-2.5, font=2)

```